---
title: "Final Project Report - Practical Machine Learning Course"
output:
  html_document:
    toc: yes
---

Description 
---
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data 
---
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

Goal
---
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

Data Preprocessing
---
First we clear the workspace and load the neded libraries.

```{r}
rm(list=ls())
#library(AppliedPredictiveModeling)
library(caret)
#library(rattle)
#library(rpart)
#library(rpart.plot)
library(randomForest)
```
**Data loading **
The data is dowloaded from internet. 
```{r}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}

```

```{r}
pmltrainingRaw <- read.table("./data/pml-training.csv",header=TRUE,sep=",")
pmltestingRaw <- read.table("./data/pml-testing.csv",header=TRUE,sep=",")
dim(pmltrainingRaw)
dim(pmltestingRaw)
craw <- colnames(pmltrainingRaw)
```
**Cleaning the data**
First we will clean the data and get rid of observations with missing values as well as some meaningless variables.
```{r}
sum(complete.cases(pmltrainingRaw))
```
We remove all the collumns with missing values.
```{r}
pmltrainingRaw <- pmltrainingRaw[, colSums(is.na(pmltrainingRaw)) == 0] 
```
We also remove the columns which have  NearZeroVariancecolumns and do not contribute much to the accelerometer measurements.

```{r}
nzv <- nearZeroVar(pmltrainingRaw, saveMetrics=TRUE)
TrainingDC <- pmltrainingRaw[,nzv$nzv==FALSE]
dim(TrainingDC)
ctraining <- colnames(TrainingDC)
cremoving <- colnames(pmltestingRaw)[!(craw %in% ctraining)]
# Remove for the testing the same columns we removed from the training 
TestingDC <- subset(pmltestingRaw,select= - which(!(craw %in% ctraining)))

```
Now we can split the training data for training (about 60 %) and validation (about 40%).
```{r}
TrainIndex <- createDataPartition(TrainingDC$classe, p=0.6, list=FALSE)
TrainingData <- TrainingDC[TrainIndex, ]
ValidationData <- TrainingDC[-TrainIndex, ]
```
Data Modeling
---
We use a Random Forest model. We fit the model on our training data, and instruct the train function to use 5-fold cross-validation to select optimal tuning parameters for the model.


```{r}
#set.seed(12345)
# instruct train to use 5-fold CV to select optimal tuning parameters
fitControl <- trainControl(method="cv", number=5)

# fit model on training Data
fit <- train(classe ~ ., data=TrainingData, method="rf", trControl=fitControl,ntree=50)
# print final model to see tuning parameters it chose
fit$finalModel
```
Now we can test the fitted model on the validation data.
```{r}
predictRF <- predict(fit, ValidationData)
confusionMatrix(ValidationData$classe, predictRF)
```
We compute the accuracy.
```{r}
accuracy <- postResample(predictRF, ValidationData$classe)
accuracy
oose <- 1 - as.numeric(confusionMatrix(ValidationData$classe, predictRF)$overall[1])
oose
```
the estimated accuracy of the model is larger tha 99% and the estimated out-of-sample error is smaller than 1%. Next we will apply the model to the testing data set.

Predicting Results on the Test Data
---
We can test now our model by trying to predict the data contained in the testing.
```{r}
#fitControl <- trainControl(method="cv", number=5)
# fit model on training Data
#fit <- train(classe ~ ., data=TrainingDC, method="rf", trControl=fitControl,ntree=50)

nresult <- predict(fit, ValidationData[2200:2250,-length(names(ValidationData))])
summary(nresult)
summary(ValidationData[2200:2250,length(names(ValidationData))])
result <- predict(fit, TestingDC)
summary(result)

if (!file.exists("./answers")) {
  dir.create("./answers")
}
pml_write_files = function(x){
  n = length(x)
  path <- "answers"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(result)